---
title: "AmbientFilter: Per-Cell-Type Ambient RNA Correction"
subtitle: "Step-by-step vignette for human islet scRNA-seq datasets"
author: "Verchere Lab"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: kate
    code_folding: show
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6,
  cache   = FALSE
)
```

---

# Background

Single-cell RNA-seq of human pancreatic islets suffers from a well-documented
ambient RNA problem: beta cells and alpha cells are so transcriptionally
dominant that their transcripts — particularly **INS**, **TTR**, and **GCG** —
contaminate the droplet soup and appear in cell types that do not actually
express them. Macrophages, lymphocytes, ductal cells, and mast cells all show
spurious islet hormone signal as a result.

`AmbientFilter` addresses this with a **hybrid two-layer approach**:

- **Layer 1 — per-sample correction**: estimates a background profile for each
  donor sample (stratified by tissue) and applies
  `DropletUtils::removeAmbience()` with data-driven or manually specified
  correction strengths.
- **Layer 2 — cross-sample consistency analysis**: identifies genes that are
  *consistently* corrected across donors — the signature of true ambient
  contamination rather than sample-specific processing noise. Produces a
  publishable ambient offenders table for the field.

Key features added since the original dRopt approach:

- **Data-driven correction strengths** via `estimate_correction_strength()`:
  learns per-cell-type correction strengths from cross-sample Spearman
  correlations. Macrophages automatically receive conservative correction
  because their genuine markers are uncorrelated with the INS/TTR ambient
  profile.
- **Prior blending**: biological knowledge (e.g. macrophages are nibbly) can
  be blended with the data-driven estimate via `prior_strengths` and
  `prior_weight`.
- **Protected genes**: genes excluded from correction entirely, preserving
  their original counts. Critical for cases where a gene may be biologically
  present via a non-ambient route — e.g. mitochondrial genes in macrophages
  that may reflect genuine beta-to-macrophage mitochondrial transfer.
- **Per-gene ambient weights**: a weight in [0,1] per gene per cell type,
  reflecting how consistently each gene tracks the background across donors.
  Used as soft supervision priors for the VAE method.
- **VAE-based correction**: a disentangled variational autoencoder that learns
  separate latent spaces for genuine expression and ambient contamination,
  available for benchmarking against the correlation method.

> **Run upstream of any subclustering.** Correct the full annotated object
> first, then extract your cell type of interest for subclustering.

---

# 1. Installation

```{r install, eval=FALSE}
# Bioconductor dependency
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("DropletUtils")

# Install AmbientFilter from GitHub
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")
devtools::install_github("jvelghe/AmbientFilter")

# For VAE method only — installs Python + PyTorch via reticulate (one time)
# AmbientFilter::install_ambient_vae()
```

---

# 2. Load libraries

```{r libraries}
library(AmbientFilter)
library(Seurat)
library(SeuratObject)
library(ggplot2)
library(patchwork)
library(dplyr)
```

---

# 3. Load your Seurat object

Your object should already be:

- QC-filtered, preprocessed (normalised, clustered), and annotated with cell
  type labels
- Containing **raw counts** in the `RNA` assay counts layer — do not use
  SCTransform-corrected counts; use the original RNA counts
- Have a sample/donor ID column and ideally a tissue column in metadata if you
  have multiple tissues (pancreas + spleen, etc.)

```{r load-object, eval=FALSE}
# Check if object is already in environment before loading
if (!exists("seu") || !is(seu, "Seurat")) {
  seu <- readRDS("~/path/to/your/Islet_subclustered.rds")
}
```

```{r inspect-object}
# Dimensions
dim(seu)

# All metadata columns
colnames(seu@meta.data)

# Cell type labels — exact spelling matters for the pipeline
table(seu@meta.data$cell_type)    # replace with your cell_type column name

# Sample IDs
table(seu@meta.data$Sample)       # replace with your sample column name

# Confirm raw counts are accessible (Seurat v5: join layers if needed)
tryCatch(
  SeuratObject::LayerData(seu, assay = "RNA", layer = "counts")[1:3, 1:3],
  error = function(e) {
    message("Joining layers (Seurat v5)...")
    seu <<- SeuratObject::JoinLayers(seu)
  }
)
```

---

# 4. Configure your run

Set column names and cell type lists here once — everything below uses these
variables.

```{r configure}
# ── Column names (match your metadata exactly) ────────────────────────────────
CELL_TYPE_COL <- "cell_type"   # column with cell type labels
SAMPLE_COL    <- "Sample"      # column with donor/sample IDs
TISSUE_COL    <- "Tissue"      # tissue column; set to NULL if single tissue
ASSAY         <- "RNA"         # assay containing raw counts

# ── Cell types to correct ─────────────────────────────────────────────────────
# Exclude endocrine types — they ARE the ambient source and correcting them
# removes real signal.
cell_types_to_correct <- c(
  "Macrophages",
  "Lymphocytes",
  "Mast",
  "Ductal",
  "Activated_Stellate",
  "Quiescent_Stellate",
  "Endothelial",
  "Fibroblasts",
  "Pericytes",
  "Schwann",
  "B_Cells"
  # Consider adding: "Acinar" (exocrine, will have ambient INS/TTR)
  # Consider adding: "Neuronal", "Neuroendo" (NIBI-axis; use conservative prior)
)

# ── Cell types to exclude from background estimation ─────────────────────────
# These dominate the ambient pool — including them inflates the profile
exclude_from_background <- c(
  "Beta", "Alpha", "Alpha-Beta", "Delta", "Gamma", "Epsilon"
)

# ── Ambient genes to highlight in plots ───────────────────────────────────────
islet_ambient_genes <- c("INS", "TTR", "GCG", "PRSS1", "CELA3A",
                          "CLPS", "PNLIP", "REG1A")

# ── Protected genes (optional — default NULL = correction applied to all genes)
# Genes excluded from correction; original counts are preserved in the output.
# Use when a gene may be biologically present via a non-ambient route.
#
# Example: mitochondrial genes in macrophages may reflect genuine
# beta-to-macrophage mitochondrial transfer, not ambient contamination.
# Protecting them preserves this biological signal.
#
# See section 11 for full discussion and a before/after comparison.

mt_genes <- grep("^MT-", rownames(seu), value = TRUE)

protected_genes <- NULL          # default: no protection — safest starting point

# Uncomment ONE of the following to enable MT gene protection:
# protected_genes <- list(Macrophages = mt_genes)   # macrophages only
# protected_genes <- list(all = mt_genes)           # all cell types
```

---

# 5. Check ambient signal before correction

```{r pre-correction-check}
ambient_genes_present <- islet_ambient_genes[islet_ambient_genes %in% rownames(seu)]
cat("Ambient genes found in object:", paste(ambient_genes_present, collapse=", "), "\n")

# If percent.ins is pre-computed, use it as a quick ambient check
if ("percent.ins" %in% colnames(seu@meta.data)) {
  cat("\nMean percent.ins by cell type:\n")
  sort(tapply(seu@meta.data$percent.ins,
              seu@meta.data[[CELL_TYPE_COL]], mean, na.rm = TRUE),
       decreasing = TRUE) |> round(3) |> print()
}
```

```{r pre-correction-violin, fig.height=5}
VlnPlot(seu, features = ambient_genes_present, group.by = CELL_TYPE_COL,
        pt.size = 0, ncol = length(ambient_genes_present)) +
  plot_annotation(
    title    = "Ambient gene expression BEFORE correction",
    subtitle = "Signal in non-endocrine populations = contamination to remove"
  )
```

```{r pre-correction-feature}
FeaturePlot(seu, features = ambient_genes_present, ncol = 3, order = TRUE)
```

---

# 6. Estimate per-sample background profiles

```{r estimate-background}
bg <- estimate_background(
  object             = seu,
  sample_col         = SAMPLE_COL,
  tissue_col         = TISSUE_COL,
  assay              = ASSAY,
  exclude_cell_types = exclude_from_background,
  cell_type_col      = CELL_TYPE_COL,
  verbose            = TRUE
)

cat("\nBackground profiles created:", length(bg), "\n")
cat("Top 20 ambient genes —", names(bg)[1], ":\n")
head(sort(bg[[1]], decreasing = TRUE), 20)
```

```{r plot-background}
top_ambient_df <- do.call(rbind, lapply(names(bg), function(s) {
  top <- head(sort(bg[[s]], decreasing = TRUE), 15)
  data.frame(sample = s, gene = names(top),
             proportion = as.numeric(top), stringsAsFactors = FALSE)
}))

ggplot(top_ambient_df,
       aes(x = reorder(gene, proportion), y = proportion,
           fill = gene %in% c("INS", "TTR"))) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("grey70", "#C94040")) +
  coord_flip() +
  facet_wrap(~sample, scales = "free_x") +
  labs(title = "Top ambient genes per sample",
       subtitle = "Red = INS or TTR (expected to dominate in islet preparations)",
       x = NULL, y = "Proportion of ambient reads") +
  theme_minimal(base_size = 10)
```

---

# 7. Data-driven correction strength estimation

`estimate_correction_strength()` learns appropriate correction strengths from
your data by computing the cross-sample Spearman correlation between each cell
type's mean expression and the ambient background profile. Cell types whose
expression tracks the background closely get aggressive correction; those whose
expression is independent of the background get conservative correction.

**Macrophages receive lower estimated strength automatically** — their genuine
marker genes (TREM2, FOLR2, ADRA2B) are uncorrelated with INS/TTR, diluting
the signal. You can blend the data estimate with a biological prior.

```{r estimate-strength}
strength_est <- estimate_correction_strength(
  object               = seu,
  background           = bg,
  cell_type_col        = CELL_TYPE_COL,
  sample_col           = SAMPLE_COL,
  tissue_col           = TISSUE_COL,
  assay                = ASSAY,
  cell_types           = cell_types_to_correct,
  min_cells            = 10,
  min_samples          = 3,
  top_genes_pct        = 0.3,
  compute_gene_weights = TRUE,    # needed for VAE method
  prior_strengths      = c(Macrophages = 0.5),  # biological prior
  prior_weight         = 0.3,    # 30% prior, 70% data-driven
  make_plots           = TRUE,
  verbose              = TRUE
)

# Summary table
strength_est$summary
```

```{r strength-bar, fig.height=5}
# Green = conservative | Red = aggressive
# Macrophages should sit toward the conservative end
strength_est$plots$strength_bar
```

```{r strength-violin, fig.height=5}
# Per-sample correlation violin per cell type
# Wide violins = noisy estimate — consider a stronger prior for that cell type
strength_est$plots$correlation_violin
```

```{r strength-scatter, fig.height=5}
# Points on diagonal = fully data-driven
# Points off diagonal = prior blending effect
strength_est$plots$correlation_vs_strength
```

```{r gene-weights}
# Per-gene ambient weights for macrophages
# High = consistently tracks background across donors = likely ambient
# Low = independent of background = likely genuinely expressed
mac_weights <- strength_est$gene_weights$Macrophages
cat("Top 20 most ambient genes in macrophages:\n")
head(sort(mac_weights, decreasing = TRUE), 20)

cat("\nTop 20 most genuinely expressed genes in macrophages:\n")
head(sort(mac_weights, decreasing = FALSE), 20)
```

---

# 8. Run the full hybrid pipeline

`ambient_filter()` orchestrates all steps in sequence:

1. Per-sample background estimation
2. Data-driven correction strength estimation
3. Per-cell-type `removeAmbience()` correction, respecting `protected_genes`
4. Cross-sample consistency analysis (hybrid layer 2)
5. Export of shareable ambient offenders table

```{r run-ambient-filter}
results <- ambient_filter(
  object                     = seu,
  cell_type_col              = CELL_TYPE_COL,
  sample_col                 = SAMPLE_COL,
  tissue_col                 = TISSUE_COL,
  cell_types                 = cell_types_to_correct,
  assay                      = ASSAY,

  # ── Correction strength ────────────────────────────────────────────────────
  # Recommended: data-driven with optional biological prior
  auto_estimate_strength     = TRUE,
  prior_strengths            = c(Macrophages = 0.5),
  prior_weight               = 0.3,
  top_genes_pct              = 0.3,
  min_samples_for_estimation = 3,

  # Alternative: fully manual (set auto_estimate_strength = FALSE)
  # correction_strength = c(Macrophages = 0.5, Lymphocytes = 1.0, ...),
  # auto_estimate_strength = FALSE,

  # ── Background ────────────────────────────────────────────────────────────
  exclude_from_background    = exclude_from_background,

  # ── Protected genes ───────────────────────────────────────────────────────
  # NULL = no protection (default; safest starting point)
  # Set in section 4 — see section 11 for full discussion
  protected_genes            = protected_genes,

  # ── Cross-sample consistency (hybrid layer 2) ─────────────────────────────
  run_consistency_analysis   = TRUE,
  min_sample_fraction        = 0.5,   # corrected in >= 50% of donors
  min_cell_types             = 2,     # ambient in >= 2 cell types

  # ── Ambient offenders export ──────────────────────────────────────────────
  export_offenders           = TRUE,
  dataset_name               = "SS2026_islet_atlas",
  tissue                     = "human_pancreatic_islet",
  n_donors                   = length(unique(seu@meta.data[[SAMPLE_COL]])),
  platform                   = "10x_Chromium_v3",

  # ── Output settings ───────────────────────────────────────────────────────
  min_cells                  = 10,
  make_plots                 = TRUE,
  label_difference           = 0.5,
  highlight_genes            = islet_ambient_genes,
  return_full_corrected_obj  = FALSE,
  save_dir                   = "ambient_filter_results",
  verbose                    = TRUE
)
```

---

# 9. Review the summary

```{r summary-table}
results$summary
```

```{r summary-plot, fig.height=5}
ggplot(results$summary,
       aes(x = reorder(cell_type, n_genes_changed),
           y = n_genes_changed, fill = cell_type)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title    = "Genes significantly changed by ambient correction",
       subtitle = "|log2FC| > log2(1.5) between original and corrected",
       x = NULL, y = "Number of changed genes") +
  theme_minimal(base_size = 11)
```

---

# 10. Correction diagnostics per cell type

## 10a. Macrophages

```{r mac-correlation, fig.height=5}
results$plots$Macrophages$correlation
```

```{r mac-bar, fig.height=5}
results$plots$Macrophages$top_genes_bar
```

```{r mac-violin, fig.height=5}
results$plots$Macrophages$violin_original +
  plot_annotation(title = "Macrophages — BEFORE correction")
results$plots$Macrophages$violin_corrected +
  plot_annotation(title = "Macrophages — AFTER correction")
```

```{r mac-changed}
head(results$cell_type_results$Macrophages$changed_genes, 20)
```

## 10b. Lymphocytes

```{r lymph-violin, fig.height=5}
results$plots$Lymphocytes$violin_original +
  plot_annotation(title = "Lymphocytes — BEFORE correction")
results$plots$Lymphocytes$violin_corrected +
  plot_annotation(title = "Lymphocytes — AFTER correction")
```

## 10c. INS reduction across all cell types — key sanity check

```{r ins-comparison, fig.height=7}
ins_df <- do.call(rbind, lapply(names(results$cell_type_results), function(ct) {
  r <- results$cell_type_results[[ct]]
  if (!"INS" %in% rownames(r$original)) return(NULL)
  rbind(
    data.frame(cell_type  = ct, state = "Before",
               expression = as.numeric(SeuratObject::LayerData(
                 r$original, layer = "counts")["INS", ])),
    data.frame(cell_type  = ct, state = "After",
               expression = as.numeric(SeuratObject::LayerData(
                 r$corrected, layer = "counts")["INS", ]))
  )
}))
ins_df$state <- factor(ins_df$state, levels = c("Before", "After"))

ggplot(ins_df, aes(x = state, y = log1p(expression), fill = state)) +
  geom_violin(scale = "width", trim = TRUE, linewidth = 0.3) +
  geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.3, linewidth = 0.3) +
  scale_fill_manual(values = c("Before" = "#C94040", "After" = "#5FAD6E")) +
  facet_wrap(~cell_type, scales = "free_y", ncol = 4) +
  labs(title    = "INS expression before and after ambient correction",
       subtitle = "Should decrease in all non-endocrine populations",
       x = NULL, y = "INS (log1p counts)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

---

# 11. Protected genes — mitochondrial transfer

If there is biological evidence that a gene may be present in a cell type via
a non-ambient route, it should be **protected from correction** so its original
counts are preserved in the output.

The clearest case in islet biology is **beta-to-macrophage mitochondrial
transfer**: mitochondrial genes (MT-CO1, MT-ND1, MT-ATP6, etc.) in macrophages
could reflect genuine foreign mitochondria rather than ambient contamination.
Correcting them would remove a real biological signal.

**Default behaviour is `protected_genes = NULL`** — no genes are protected and
correction is applied uniformly. This is the safest starting point. Enable
protection only when you have biological evidence to support it (e.g. a bulk
RNA-seq or imaging dataset demonstrating transfer).

```{r check-mt-genes}
# How many MT genes are in this object?
mt_present <- grep("^MT-", rownames(seu), value = TRUE)
cat("Mitochondrial genes present:", length(mt_present), "\n")
cat(paste(mt_present, collapse = ", "), "\n")
```

```{r mt-before-after}
# Compare MT gene expression in macrophages before and after correction
mac_r <- results$cell_type_results$Macrophages
mt_compare <- do.call(rbind, lapply(mt_present, function(g) {
  if (!g %in% rownames(mac_r$original)) return(NULL)
  orig <- mean(as.numeric(
    SeuratObject::LayerData(mac_r$original, layer = "counts")[g, ]))
  corr <- mean(as.numeric(
    SeuratObject::LayerData(mac_r$corrected, layer = "counts")[g, ]))
  data.frame(
    gene           = g,
    mean_original  = round(orig, 3),
    mean_corrected = round(corr, 3),
    pct_change     = round((corr - orig) / (orig + 0.001) * 100, 1),
    protected      = g %in% unlist(protected_genes)
  )
}))
print(mt_compare)
# protected = FALSE and pct_change negative = corrected normally
# protected = TRUE  and pct_change ~0      = original counts restored
```

```{r mt-protected-rerun, eval=FALSE}
# To re-run WITH mitochondrial protection in macrophages:
results_protected <- ambient_filter(
  object                    = seu,
  cell_type_col             = CELL_TYPE_COL,
  sample_col                = SAMPLE_COL,
  tissue_col                = TISSUE_COL,
  cell_types                = cell_types_to_correct,
  auto_estimate_strength    = TRUE,
  prior_strengths           = c(Macrophages = 0.5),
  prior_weight              = 0.3,
  exclude_from_background   = exclude_from_background,
  protected_genes           = list(Macrophages = mt_genes),
  run_consistency_analysis  = TRUE,
  export_offenders          = TRUE,
  dataset_name              = "SS2026_islet_atlas_mt_protected",
  save_dir                  = "ambient_filter_results_protected",
  verbose                   = TRUE
)
```

---

# 12. Cross-sample consistency — the hybrid layer

```{r consistency-heatmap, fig.height=10, fig.width=10}
# Heatmap: genes x cell types
# Colour = fraction of donors in which this gene was corrected
# Dark green = consistently ambient across most donors
results$ambient_map$plots$heatmap
```

```{r cross-celltype-bar, fig.height=6}
# Genes ambient in >= min_cell_types cell types — the strongest offenders
results$ambient_map$plots$cross_cell_type_bar
```

```{r consistency-dot, fig.height=8}
# Dot plot: per-cell-type consistency
# Size = fraction of donors | Colour = mean log2FC
results$ambient_map$plots$per_sample_consistency
```

```{r offenders-table}
# Cross-cell-type ambient offenders — your publishable list
results$ambient_map$cross_cell_type
```

```{r recommended-features}
# Consistently ambient genes per cell type
# Feed into a second-pass correction to force-correct these genes
results$ambient_map$recommended_features
```

---

# 13. Optional: two-pass correction

Run a second pass using `features = recommended_features` to force-correct
consistently ambient genes that may have been missed in samples with very few
cells of a given type.

```{r two-pass, eval=FALSE}
results_v2 <- ambient_filter(
  object                    = seu,
  cell_type_col             = CELL_TYPE_COL,
  sample_col                = SAMPLE_COL,
  tissue_col                = TISSUE_COL,
  cell_types                = cell_types_to_correct,
  auto_estimate_strength    = TRUE,
  prior_strengths           = c(Macrophages = 0.5),
  prior_weight              = 0.3,
  exclude_from_background   = exclude_from_background,
  protected_genes           = protected_genes,
  features                  = results$ambient_map$recommended_features,
  run_consistency_analysis  = TRUE,
  export_offenders          = TRUE,
  dataset_name              = "SS2026_islet_atlas_v2",
  save_dir                  = "ambient_filter_results_v2",
  verbose                   = TRUE
)
```

---

# 14. Ambient offenders — shareable community resource

The ambient offenders table is saved to
`ambient_filter_results/ambient_offenders/` as CSV, RDS, and Excel.
Sharing this with the field helps other islet labs clean their own data.

```{r offenders-export, eval=FALSE}
# Export manually with custom metadata
export_ambient_offenders(
  ambient_map  = results$ambient_map,
  dataset_name = "SS2026_islet_atlas_2026",
  tissue       = "human_pancreatic_islet",
  n_donors     = length(unique(seu@meta.data[[SAMPLE_COL]])),
  platform     = "10x_Chromium_v3",
  notes        = "CD45-enriched + whole pancreas; SS2026 JDRF2 atlas",
  save_dir     = "ambient_filter_results/ambient_offenders"
)
```

```{r offenders-load, eval=FALSE}
# How another lab uses your offenders file
offenders <- readRDS(
  "ambient_filter_results/ambient_offenders/ambient_offenders.rds"
)
offenders$metadata   # provenance: dataset, tissue, platform, n_donors
offenders$offenders  # the gene list
```

---

# 15. Extract corrected objects for subclustering

```{r extract-corrected}
# All original metadata is preserved in corrected objects
mac_corrected  <- results$cell_type_results$Macrophages$corrected
lymp_corrected <- results$cell_type_results$Lymphocytes$corrected
mast_corrected <- results$cell_type_results$Mast$corrected

# Verify metadata survived
cat("Columns in corrected macrophage object:\n")
print(colnames(mac_corrected@meta.data))
```

```{r subcluster-pipeline, eval=FALSE}
# Standard Seurat subclustering — same pipeline as before correction
mac_corrected <- mac_corrected |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA() |>
  IntegrateLayers(
    method         = SeuratWrappers::FastMNNIntegration,
    new.reduction  = "integrated.mnn",
    orig.reduction = "pca"
  ) |>
  FindNeighbors(reduction = "integrated.mnn", dims = 1:30) |>
  FindClusters(resolution = 0.3) |>
  RunUMAP(reduction = "integrated.mnn", dims = 1:30)

DimPlot(mac_corrected, group.by = "seurat_clusters", label = TRUE)
DimPlot(mac_corrected, group.by = "Condition")
```

---

# 16. Optional: rebuild full corrected Seurat object

```{r full-object, eval=FALSE}
# Memory-intensive — duplicates the full object
results_full <- ambient_filter(
  object                    = seu,
  cell_type_col             = CELL_TYPE_COL,
  sample_col                = SAMPLE_COL,
  tissue_col                = TISSUE_COL,
  cell_types                = cell_types_to_correct,
  auto_estimate_strength    = TRUE,
  prior_strengths           = c(Macrophages = 0.5),
  prior_weight              = 0.3,
  exclude_from_background   = exclude_from_background,
  protected_genes           = protected_genes,
  return_full_corrected_obj = TRUE,
  save_dir                  = "ambient_filter_results",
  verbose                   = TRUE
)

seu_corrected <- results_full$updated_object
saveRDS(seu_corrected, "Islet_subclustered_ambient_corrected.rds")
```

---

# 17. Optional: VAE-based correction

The VAE method trains a disentangled variational autoencoder with separate
latent spaces for genuine expression (`z_bio`) and ambient contamination
(`z_ambient`). Gene weights from `estimate_correction_strength()` serve as
soft supervision priors. Use for benchmarking or final publication-ready
correction.

```{r vae-install, eval=FALSE}
# One-time Python + PyTorch installation
AmbientFilter::install_ambient_vae()
```

```{r vae-run, eval=FALSE}
results_vae <- ambient_filter_vae(
  object             = seu,
  background         = bg,
  cell_types         = cell_types_to_correct,
  cell_type_col      = CELL_TYPE_COL,
  sample_col         = SAMPLE_COL,
  tissue_col         = TISSUE_COL,
  strength_estimates = strength_est,    # gene weights as VAE priors
  protected_genes    = protected_genes, # same MT protection applies
  n_epochs           = 150L,
  latent_dim_bio     = 10L,
  latent_dim_ambient = 3L,
  save_dir           = "ambient_filter_vae_results",
  verbose            = TRUE
)

# Training loss curve + per-cell ambient scores
plot_vae_diagnostics(results_vae$cell_type_results$Macrophages,
                     cell_label = "Macrophages")
```

```{r vae-benchmark, eval=FALSE}
# Side-by-side comparison of correlation vs VAE
bench <- benchmark_methods(results, results_vae,
                           highlight_genes = islet_ambient_genes)
bench$benchmark_plot
bench$comparison_table
```

---

# 18. Recommended: compare DE before and after correction

```{r de-comparison, eval=FALSE}
library(muscat)
library(DESeq2)

run_pseudobulk_de <- function(obj, sample_col, group_col, contrast) {
  sce <- as.SingleCellExperiment(obj)
  sce <- prepSCE(sce, kid = CELL_TYPE_COL, gid = group_col,
                 sid = sample_col, drop = TRUE)
  pb  <- aggregateData(sce, assay = "counts", fun = "sum")
  res <- pbDS(pb, design = ~group_id, contrast = contrast, verbose = FALSE)
  resDS(sce, res, bind = "row", frq = TRUE)
}

de_original  <- run_pseudobulk_de(
  results$cell_type_results$Macrophages$original,
  sample_col = SAMPLE_COL, group_col = "Condition", contrast = "T1D - ND"
)
de_corrected <- run_pseudobulk_de(
  results$cell_type_results$Macrophages$corrected,
  sample_col = SAMPLE_COL, group_col = "Condition", contrast = "T1D - ND"
)

top_orig <- head(de_original[order(de_original$p_adj.loc),  "gene"], 50)
top_corr <- head(de_corrected[order(de_corrected$p_adj.loc), "gene"], 50)

cat("Likely ambient false positives (in original, gone after correction):\n")
print(setdiff(top_orig, top_corr))

cat("\nUnmasked by correction (absent in original, appear after correction):\n")
print(setdiff(top_corr, top_orig))
```

---

# 19. Session info

```{r session-info}
sessionInfo()
```
