---
title: "AmbientFilter: Per-Cell-Type Ambient RNA Correction"
subtitle: "Step-by-step vignette for human islet scRNA-seq datasets"
author: "Verchere Lab"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: kate
    code_folding: show
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6,
  cache   = FALSE
)
```

---

# Background

Single-cell RNA-seq of human pancreatic islets suffers from a well-documented
ambient RNA problem: beta cells and alpha cells are so transcriptionally
dominant that their transcripts — particularly **INS**, **TTR**, and **GCG** —
contaminate the droplet soup and appear in cell types that don't actually
express them. Macrophages, lymphocytes, ductal cells, and mast cells all show
spurious islet hormone signal as a result.

This vignette walks through `AmbientFilter`, which extends the approach of
[dRopt](https://github.com/CUAnschutzBDC/dRopt) to correct ambient
contamination on a **per-cell-type basis**, with cell-type-specific correction
strengths. Crucially, macrophages receive a **conservative correction** (0.5)
because they legitimately internalise ambient RNA via efferocytosis and
trogocytosis — being "nibbly" is part of their biology.

> **Run this upstream of any subclustering.** Correct the full annotated object
> first, then extract your cell type of interest for subclustering.

---

# 1. Installation

```{r install, eval=FALSE}
# Bioconductor dependency
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("DropletUtils")

# Install AmbientFilter from GitHub
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")
devtools::install_github("jvelghe/AmbientFilter")
```

---

# 2. Load libraries

```{r libraries}
library(AmbientFilter)
library(Seurat)
library(SeuratObject)
library(ggplot2)
library(patchwork)
library(dplyr)
```

---

# 3. Load your Seurat object

Your object should already be:

- QC-filtered and preprocessed (normalised, clustered)
- Annotated with cell type labels in a metadata column
- Containing **raw counts** in the `RNA` assay counts layer (do not use
  SCTransform-corrected counts as input here — use the original RNA counts)
- Split by sample in a metadata column (one value per donor)
- Ideally, tissue of origin in a separate metadata column if you have
  pancreas + spleen + other tissues in one object

```{r load-object, eval=FALSE}
# Load your object
seu <- readRDS("/path/to/your/islet_atlas.rds")
```

```{r inspect-object}
# For this vignette, we'll assume `seu` is already loaded.
# Inspect the key metadata columns:
head(seu@meta.data[, c("orig.ident",   # or your sample column
                        "tissue",        # or your tissue column
                        "cell_type")])   # or your cell type column

# Check what cell types are present
table(seu@meta.data$cell_type)

# Check what samples are present
table(seu@meta.data$orig.ident)

# Make sure raw counts are in the RNA assay
GetAssayData(seu, assay = "RNA", layer = "counts")[1:5, 1:5]
```

---

# 4. Check ambient signal before correction

Before running the correction, let's visualise the ambient signal we expect to
remove. In islet data, INS and TTR should ideally be near-zero in
non-endocrine populations.

```{r pre-correction-violin}
# Canonical ambient genes in islet scRNA-seq
ambient_genes <- c("INS", "TTR", "GCG", "PRSS1", "CELA3A")

# Only plot genes that are present in your object
ambient_genes_present <- ambient_genes[ambient_genes %in% rownames(seu)]

# Violin plot split by cell type — look for signal in non-expressing populations
VlnPlot(
  seu,
  features  = ambient_genes_present,
  group.by  = "cell_type",
  pt.size   = 0,
  ncol      = length(ambient_genes_present)
) +
  plot_annotation(
    title    = "Ambient gene expression BEFORE correction",
    subtitle = "INS/TTR/GCG should be near-zero in macrophages, lymphocytes, ductal, mast cells"
  )
```

```{r pre-correction-feature}
# Feature plots to see spatial distribution of ambient signal
FeaturePlot(
  seu,
  features = ambient_genes_present,
  ncol     = 3,
  order    = TRUE
)
```

---

# 5. Estimate per-sample background profiles

The background profile represents the expected ambient distribution — the
proportion of total reads mapping to each gene across all cells in a sample.
In islet data, INS and TTR will dominate this profile.

Key decisions here:

- **`tissue_col`**: always specify this if you have multiple tissues. A spleen
  background is very different from a pancreas background, and mixing them will
  over- or under-correct.
- **`exclude_from_background`**: if your samples are heavily enriched for islet
  cells (e.g. CD45-depleted or islet-enriched preps), you may want to exclude
  beta and alpha cells from the background estimate to avoid them dominating the
  ambient profile to an unrealistic degree. For whole-pancreas dissociations,
  leave this as `NULL`.

```{r estimate-background}
bg <- estimate_background(
  object             = seu,
  sample_col         = "orig.ident",   # <-- change to your sample column
  tissue_col         = "tissue",        # <-- change to your tissue column, or NULL
  assay              = "RNA",
  exclude_cell_types = NULL,            # e.g. c("beta", "alpha") if enriched
  cell_type_col      = "cell_type",
  verbose            = TRUE
)

# Inspect: what are the top ambient genes for one sample?
sample_id <- names(bg)[1]
cat("\nTop 20 ambient genes for sample:", sample_id, "\n")
head(sort(bg[[sample_id]], decreasing = TRUE), 20)
```

```{r plot-background}
# Plot the ambient profile for each sample — INS/TTR should dominate
# in pancreas samples
top_ambient <- lapply(names(bg), function(s) {
  top <- head(sort(bg[[s]], decreasing = TRUE), 15)
  data.frame(
    sample     = s,
    gene       = names(top),
    proportion = as.numeric(top),
    stringsAsFactors = FALSE
  )
})
top_ambient_df <- do.call(rbind, top_ambient)

ggplot(top_ambient_df, aes(x = reorder(gene, proportion), y = proportion,
                            fill = gene == "INS" | gene == "TTR")) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("grey70", "#C94040")) +
  coord_flip() +
  facet_wrap(~sample, scales = "free_x") +
  labs(
    title    = "Top ambient genes per sample",
    subtitle = "Red = INS or TTR (expected to dominate in islet preparations)",
    x        = NULL,
    y        = "Proportion of all reads"
  ) +
  theme_minimal(base_size = 10)
```

---

# 6. Run the full ambient correction pipeline

Now we run the main function. Define your correction strengths carefully:

| Cell type | Strength | Reasoning |
|-----------|----------|-----------|
| `macrophage` | `0.5` | Legitimately nibbly — efferocytosis/trogocytosis |
| `lymphocyte` | `1.0` | No biological reason to contain ambient transcripts |
| `mast_cell` | `1.0` | No biological reason |
| `ductal` | `1.0` | No biological reason |
| `stellate` | `1.0` | No biological reason |
| `endothelial` | `1.0` | No biological reason |
| `beta`, `alpha` | skip or `0.0` | They *are* the ambient source |

```{r define-parameters}
# Define which cell types to correct
# Exclude endocrine cells — they ARE the ambient source
cell_types_to_correct <- c(
  "macrophage",
  "lymphocyte",
  "mast_cell",
  "ductal",
  "stellate",
  "endothelial",
  "schwann"
  # Add any other non-endocrine cell types in your atlas
)

# Per-cell-type correction strength
# Adjust this named vector to match your cell_type labels exactly
correction_strengths <- c(
  macrophage  = 0.5,   # conservative — macrophages are nibbly
  lymphocyte  = 1.0,
  mast_cell   = 1.0,
  ductal      = 1.0,
  stellate    = 1.0,
  endothelial = 1.0,
  schwann     = 1.0
)

# Islet-specific genes to always highlight in diagnostic plots
islet_ambient_genes <- c("INS", "TTR", "GCG", "PRSS1", "CELA3A",
                          "CLPS", "PNLIP", "REG1A")
```

```{r run-ambient-filter}
# Run the full hybrid pipeline:
#   Step 1:   per-sample background estimation
#   Step 1.5: data-driven correction strength estimation (NEW — replaces manual values)
#   Step 2:   per-cell-type removeAmbience() correction
#   Step 3:   cross-sample consistency analysis (the pooled hybrid layer)
#   Step 4:   export shareable ambient offenders table

results <- ambient_filter(
  object                    = seu,
  cell_type_col             = "cell_type",
  sample_col                = "orig.ident",
  tissue_col                = "tissue",          # NULL if single tissue
  cell_types                = cell_types_to_correct,

  # ── Data-driven correction strength (recommended) ─────────────────────────
  # auto_estimate_strength = TRUE learns correction strengths from cross-sample
  # ambient correlations — no need to manually set correction_strength.
  # Macrophages will automatically receive conservative correction because their
  # genuine marker genes are uncorrelated with the ambient background.
  auto_estimate_strength    = TRUE,

  # Optional: blend data-driven estimate with biological prior knowledge.
  # prior_weight = 0 means fully data-driven.
  # prior_weight = 1 means use prior_strengths exactly (same as manual).
  prior_strengths           = c(macrophage = 0.5),  # biological prior
  prior_weight              = 0.3,    # 30% prior, 70% data-driven

  # Alternatively: supply correction_strength manually and set
  # auto_estimate_strength = FALSE to use the old approach:
  # correction_strength = correction_strengths,
  # auto_estimate_strength = FALSE,

  exclude_from_background   = NULL,
  assay                     = "RNA",
  min_cells                 = 10,
  top_genes_pct             = 0.3,    # use top 30% expressed genes for correlation
  min_samples_for_estimation = 3,     # min samples needed for reliable estimate
  make_plots                = TRUE,
  label_difference          = 0.5,
  highlight_genes           = islet_ambient_genes,

  # ── Hybrid method: cross-sample consistency analysis ──────────────────────
  run_consistency_analysis  = TRUE,
  min_sample_fraction       = 0.5,
  min_cell_types            = 2,

  # ── Shareable ambient offenders export ────────────────────────────────────
  export_offenders          = TRUE,
  dataset_name              = "HPAP_islet_atlas_2026",
  tissue                    = "human_pancreatic_islet",
  n_donors                  = 87,
  platform                  = "10x_Chromium_v3",
  return_full_corrected_obj = FALSE,
  save_dir                  = "ambient_filter_results",
  verbose                   = TRUE
)
```

---

# 6b. Inspect data-driven correction strength estimates

Before reviewing correction results, check what strengths the model estimated.
This replaces your manual correction_strength guesses and is derived entirely
from your data.

```{r strength-estimates}
# Summary: what correction strength was estimated per cell type?
results$strength_estimates$summary
```

```{r strength-bar, fig.height=5}
# Bar chart: estimated correction strength per cell type
# Green = conservative (cell type has genuine expression independent of ambient)
# Red = aggressive (cell type expression tracks ambient background closely)
# Macrophages should automatically cluster toward the conservative end
results$strength_estimates$plots$strength_bar
```

```{r strength-violin, fig.height=5}
# Per-sample correlation violin: shows how consistent the ambient correlation
# is across donors — high variance = noisy estimate, consider providing a prior
results$strength_estimates$plots$correlation_violin
```

```{r strength-scatter, fig.height=5}
# Correlation vs final strength — shows effect of prior blending
# Points on the diagonal = fully data-driven
# Points off diagonal = pulled toward prior
results$strength_estimates$plots$correlation_vs_strength
```

```{r gene-weights}
# Per-gene ambient weights for macrophages
# High weight (near 1) = gene is consistently ambient across donors
# Low weight (near 0) = gene is genuinely expressed in macrophages
mac_weights <- results$strength_estimates$gene_weights$macrophage
# Top ambient genes in macrophages according to the model
head(sort(mac_weights, decreasing = TRUE), 20)
```

---

# 7. Review the correction summary

```{r summary}
# Overall summary across all cell types
results$summary
```

```{r summary-plot}
# Visualise: how many genes were significantly changed per cell type?
ggplot(results$summary,
       aes(x = reorder(cell_type, n_genes_changed), y = n_genes_changed,
           fill = cell_type)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title    = "Genes significantly changed by ambient correction",
    subtitle = "|log2FC| > log2(1.5) between original and corrected",
    x        = NULL,
    y        = "Number of changed genes"
  ) +
  theme_minimal(base_size = 11)
```

---

# 8. Inspect diagnostics per cell type

## 8a. Macrophages (conservative correction)

```{r mac-diagnostics, fig.height=5}
# Correlation plot: original vs corrected mean expression
# Points below the diagonal were reduced by correction
results$plots$macrophage$correlation
```

```{r mac-top-genes, fig.height=5}
# Bar chart of most-changed genes
results$plots$macrophage$top_genes_bar
```

```{r mac-violin, fig.height=5}
# Violin plots — before and after for the top changed genes
results$plots$macrophage$violin_original  +
  plot_annotation(title = "Macrophage — BEFORE correction")

results$plots$macrophage$violin_corrected +
  plot_annotation(title = "Macrophage — AFTER correction")
```

```{r mac-changed-genes}
# Table of most-changed genes in macrophages
head(results$cell_type_results$macrophage$changed_genes, 20)
```

## 8b. Lymphocytes (full correction)

```{r lymph-diagnostics, fig.height=5}
results$plots$lymphocyte$correlation
```

```{r lymph-violin, fig.height=5}
results$plots$lymphocyte$violin_original  +
  plot_annotation(title = "Lymphocyte — BEFORE correction")

results$plots$lymphocyte$violin_corrected +
  plot_annotation(title = "Lymphocyte — AFTER correction")
```

```{r lymph-changed-genes}
head(results$cell_type_results$lymphocyte$changed_genes, 20)
```

## 8c. INS expression across all corrected cell types

This is the key sanity check — INS should drop substantially in all
non-endocrine populations after correction.

```{r ins-comparison, fig.height=7}
# Compare INS expression before and after correction across all cell types
ins_before <- lapply(names(results$cell_type_results), function(ct) {
  obj  <- results$cell_type_results[[ct]]$original
  expr <- GetAssayData(obj, assay = "RNA", layer = "counts")["INS", ]
  data.frame(
    cell_type  = ct,
    expression = as.numeric(expr),
    state      = "Before",
    stringsAsFactors = FALSE
  )
})

ins_after <- lapply(names(results$cell_type_results), function(ct) {
  obj  <- results$cell_type_results[[ct]]$corrected
  if (!"INS" %in% rownames(obj)) return(NULL)
  expr <- GetAssayData(obj, assay = "RNA", layer = "counts")["INS", ]
  data.frame(
    cell_type  = ct,
    expression = as.numeric(expr),
    state      = "After",
    stringsAsFactors = FALSE
  )
})

ins_df <- do.call(rbind, c(ins_before, ins_after))
ins_df$state <- factor(ins_df$state, levels = c("Before", "After"))

ggplot(ins_df, aes(x = state, y = log1p(expression), fill = state)) +
  geom_violin(scale = "width", trim = TRUE, linewidth = 0.3) +
  geom_boxplot(width = 0.1, outlier.size = 0.3, fill = "white",
               linewidth = 0.3) +
  scale_fill_manual(values = c("Before" = "#C94040", "After" = "#5FAD6E")) +
  facet_wrap(~cell_type, scales = "free_y", ncol = 4) +
  labs(
    title    = "INS expression before and after ambient correction",
    subtitle = "log1p(counts) — should decrease in all non-endocrine populations",
    x        = NULL,
    y        = "INS expression (log1p counts)"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

---

# 9. Extract corrected objects for subclustering

Once you're satisfied with the correction, pull out the corrected subsets for
downstream subclustering. These are clean Seurat objects with corrected counts
and re-normalised data layers, ready to feed into your standard pipeline.

```{r extract-for-subclustering, eval=FALSE}
# Macrophage object — corrected, ready for subclustering
mac_corrected <- results$cell_type_results$macrophage$corrected

# Lymphocyte object
lymph_corrected <- results$cell_type_results$lymphocyte$corrected

# Mast cell object
mast_corrected <- results$cell_type_results$mast_cell$corrected

# These are standard Seurat objects — proceed with your normal pipeline
# e.g. for macrophages:
mac_corrected <- mac_corrected |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA() |>
  IntegrateLayers(
    method    = SeuratWrappers::FastMNNIntegration,
    new.reduction = "integrated.mnn",
    orig.reduction = "pca"
  ) |>
  FindNeighbors(reduction = "integrated.mnn", dims = 1:30) |>
  FindClusters(resolution = 0.3) |>
  RunUMAP(reduction = "integrated.mnn", dims = 1:30)

DimPlot(mac_corrected, group.by = "seurat_clusters", label = TRUE)
```

---

# 9b. Cross-sample consistency analysis — the hybrid layer

The hybrid method automatically runs after per-sample correction.
It asks: *which genes are consistently corrected across donors* — the
signature of true ambient contamination rather than sample-specific noise.

```{r consistency-heatmap, fig.height=10, fig.width=10}
# The ambient map is stored in results$ambient_map
# Heatmap: genes (rows) x cell types (columns)
# Colour = fraction of samples in which this gene was corrected
# Dark green = consistently ambient across most donors
results$ambient_map$plots$heatmap
```

```{r cross-celltype-offenders}
# Genes that are ambient in >= min_cell_types cell types
# These are your strongest ambient offenders
results$ambient_map$cross_cell_type
```

```{r cross-celltype-bar, fig.height=6}
# Bar chart of cross-cell-type offenders ranked by consistency
results$ambient_map$plots$cross_cell_type_bar
```

```{r per-sample-dot, fig.height=8}
# Dot plot: per-cell-type consistency
# Dot size = fraction of samples corrected | colour = mean log2FC
results$ambient_map$plots$per_sample_consistency
```

```{r recommended-features}
# Recommended features for a second-pass correction
# These are the consistently ambient genes per cell type
# You can feed this directly into a second ambient_filter() run
results$ambient_map$recommended_features
```

---

# 9c. Optional: two-pass correction with force-correction of consistent offenders

If some consistent ambient genes weren't fully removed in the first pass
(e.g. in samples with very few cells), run a second pass with
`features = recommended_features` to force-correct them.

```{r two-pass, eval=FALSE}
results_v2 <- ambient_filter(
  object              = seu,
  cell_type_col       = "cell_type",
  sample_col          = "orig.ident",
  tissue_col          = "tissue",
  cell_types          = cell_types_to_correct,
  correction_strength = correction_strengths,
  features            = results$ambient_map$recommended_features, # <-- force-correct
  run_consistency_analysis = TRUE,
  min_sample_fraction = 0.5,
  min_cell_types      = 2,
  export_offenders    = TRUE,
  dataset_name        = "HPAP_islet_atlas_2026_v2",
  save_dir            = "ambient_filter_results_v2",
  verbose             = TRUE
)
```

---

# 10. Ambient offenders — shareable community resource

The ambient offenders table is automatically saved to
`ambient_filter_results/ambient_offenders/` in three formats:

- `ambient_offenders.csv` — plain text with metadata header, shareable with anyone
- `ambient_offenders.rds` — R object for direct loading
- `ambient_offenders.xlsx` — Excel with per-cell-type sheets and colour-coded
  consistency matrix

These files identify genes that are **consistently ambient across donors and
cell types** in your specific tissue/platform combination. Sharing these with
the field helps other labs using the same tissue clean their own data.

```{r view-offenders}
# View the ambient offenders table
results$ambient_map$cross_cell_type
```

```{r load-offenders-rds, eval=FALSE}
# How another lab would load and use your offenders file:
offenders <- readRDS("ambient_filter_results/ambient_offenders/ambient_offenders.rds")

# Their top ambient genes
offenders$offenders$gene

# Use as features in their own correction run
their_results <- ambient_filter(
  their_object,
  features = list(
    macrophage = offenders$offenders$gene,
    lymphocyte = offenders$offenders$gene
  ),
  ...
)
```

```{r export-offenders-manual, eval=FALSE}
# Or export manually with custom metadata:
export_ambient_offenders(
  ambient_map  = results$ambient_map,
  dataset_name = "HPAP_islet_atlas_2026",
  tissue       = "human_pancreatic_islet",
  n_donors     = 87,
  platform     = "10x_Chromium_v3",
  notes        = "CD45-enriched + whole pancreas; SS2026 JDRF2 atlas",
  save_dir     = "ambient_filter_results/ambient_offenders"
)
```

---

# 11. Optional: rebuild the full corrected object

If you want to continue working with the full multi-cell-type object with all
corrections applied (e.g. for cross-cell-type analyses), re-run with
`return_full_corrected_obj = TRUE`. Note this is memory-intensive as it
duplicates the Seurat object.

```{r full-object, eval=FALSE}
results_full <- ambient_filter(
  object                    = seu,
  cell_type_col             = "cell_type",
  sample_col                = "orig.ident",
  tissue_col                = "tissue",
  cell_types                = cell_types_to_correct,
  correction_strength       = correction_strengths,
  highlight_genes           = islet_ambient_genes,
  return_full_corrected_obj = TRUE,   # <-- write back to full object
  save_dir                  = "ambient_filter_results",
  verbose                   = TRUE
)

# The full updated object
seu_corrected <- results_full$updated_object

# Save it
saveRDS(seu_corrected, "seu_ambient_corrected.rds")
```

---

# 12. Recommended: compare DE before and after

Following the dRopt recommendation, always run differential expression on both
the original and corrected objects and compare gene lists. In our experience,
the gene lists are similar but the corrected version no longer includes
cell-type-specific contamination genes (INS, TTR, GCG) as false positives.

```{r de-comparison, eval=FALSE}
# Example: pseudobulk DE on macrophages before and after correction
# Using muscat + DESeq2 (recommended)
library(muscat)
library(DESeq2)

# Helper to run pseudobulk DE with muscat
run_pseudobulk_de <- function(obj, sample_col, group_col, contrast) {
  sce <- as.SingleCellExperiment(obj)
  sce <- prepSCE(
    sce,
    kid = "cell_type",   # cluster column
    gid = group_col,     # group column (e.g. disease condition)
    sid = sample_col,    # sample column
    drop = TRUE
  )
  pb  <- aggregateData(sce, assay = "counts", fun = "sum")
  res <- pbDS(pb, design = ~group_id, contrast = contrast, verbose = FALSE)
  resDS(sce, res, bind = "row", frq = TRUE)
}

# Run on original macrophages
de_original <- run_pseudobulk_de(
  results$cell_type_results$macrophage$original,
  sample_col = "orig.ident",
  group_col  = "condition",  # e.g. "T1D" vs "ND"
  contrast   = "T1D - ND"
)

# Run on corrected macrophages
de_corrected <- run_pseudobulk_de(
  results$cell_type_results$macrophage$corrected,
  sample_col = "orig.ident",
  group_col  = "condition",
  contrast   = "T1D - ND"
)

# Compare: which top DE genes disappeared after correction?
top_original <- head(de_original[order(de_original$p_adj.loc), "gene"], 50)
top_corrected <- head(de_corrected[order(de_corrected$p_adj.loc), "gene"], 50)

cat("Genes in original DE but NOT in corrected (likely ambient):\n")
print(setdiff(top_original, top_corrected))

cat("\nGenes in corrected DE but NOT in original (unmasked by correction):\n")
print(setdiff(top_corrected, top_original))
```

---

# 13. Session info

```{r session-info}
sessionInfo()
```
